---
title: "Biological ontologies and machine learning"
author: Correspondence - t.konopka@qmul.ac.uk
output:
  pdf_document:
    keep_tex: true
header-includes:
  \usepackage{caption}
  \usepackage{float}
  \usepackage{amsmath}
  \usepackage{graphicx}
  \usepackage[ruled]{algorithm2e}
  \floatplacement{figure}{H}
  \captionsetup[figure]{labelfont={bf}, name={Figure}, labelsep=period}
---


```{r, analysis.configuration, echo=FALSE}
source("config.R")
source("OML_wrappers.R")
set.seed(1234)
```

```{r, ontologies, echo=FALSE}
source("OML_overview.R")
source("OML_search.R")
source("OML_graph.R")
source("OML_casestudies.R")
```

# Results

```{r fig.all, eval=TRUE, echo=FALSE, fig.width=4.8, fig.height=1.3, fig.cap="\\textbf{Summary of biological ontologies.} Panels show a collection of ontologies stratified by \\textbf{(A)} the number of terms (classes), \\textbf{(B)} the time since last update, and \\textbf{(C)} the number of disconnected components. Ontologies in light-colored groups were excluded from subsequent analysis."}
layout(matrix(c(1,2,3),	ncol=3, nrow=1, byrow=T),
       widths=c(1.6, 1.6, 1.6), height=c(1.2, 1.2, 2.7))
#
# panel explaining number of terms in ontologies
plot.custom.histogram(ontologies.summary$num_terms,
                      intervals=c("< 10"=10, "10-100"=100,
                                  "100-1K"=1000, "1K-10K"=10000,
                                  "10K-100K"=100000, "100K-1M"=1e6, "> 1M"=Inf),
                      filtered="< 10",
                      main="Size (number of terms)", ylab="Ontologies")
multipanelLabel(panel.labels[1])
#
# panel explaining how certain ontologies are excluded
plot.custom.histogram(ontologies.summary$age,
                      intervals=c("< 1 month"=30, "1-3 months"=91,
                                  "3-6 months"=182, "6-12 months"=365,
                                  "1-2 years"=730, "> 2 years"=Inf, "NA"=NA),
                      filtered=c("> 2 years", "NA"),
                      main="Time since last update", ylab="Ontologies")
multipanelLabel(panel.labels[2])
plot.custom.histogram(field.widestats$count_zero_depth,
                      intervals=c("1"=1, "2"=2, "3"=3, "4"=4, "5"=5,
		                          "6-10"=10, "> 10"=Inf),
					  filtered=c("> 10"),
                      main="Number of tree roots", ylab="Ontologies")
multipanelLabel(panel.labels[3])
```

Total number of ontologies: `r nrow(ontologies.summary)`

Number of ontologies selected for downstream presentation: `r length(ontologies.selected)`

Range of ontology sizes: `r min(ontologies.summary$num_terms)` - `r max(ontologies.summary$num_terms)`

Number older than 2 years: `r sum(ontologies.summary$age>730, na.rm=TRUE)`

Number age NA: `r sum(is.na(ontologies.summary$age))`

Number single root: `r sum(field.widestats$count_zero_depth==1, na.rm=TRUE)`

Number more than 10 roots: `r sum(field.widestats$count_zero_depth>10, na.rm=TRUE)`

Distribution of median characters in definitions:

 ```{r}
summary(field.widestats[q50_chars_def>0]$q50_chars_def)
```


```{r fig.ontologies, eval=TRUE, echo=FALSE, fig.width=6.5, fig.height=6.7, fig.cap="\\textbf{Summary statistics for selected ontologies.} Vertical panels display the number of terms in each ontology, the distribution of term depths (normalized by the size of each ontology), the distribution of the number of parents for each term, the proportion of terms lacking a text definition, the distribution of number of characters in text definitions, and the distribution of the number of synonyms. Distributions are visualized by a median (central bar), interquartile range (boxes), 5\\%-95\\% quantile range (whiskers), and maximal values (dots)."}
layout(matrix(c(1,2,3,4,5,6,7), ncol=7, nrow=1, byrow=T),
       widths=c(1.15, 0.9, 0.9, 0.9, 0.9, 0.9))
fig.widestats = field.widestats[id %in% ontologies.selected]
fig.widestats$ID = toupper(fig.widestats$id)
fig.onto.summary(fig.widestats[order(num_terms, decreasing=TRUE)])
```


```{r fig.reconstruction, eval=TRUE, echo=FALSE, fig.width=3.0, fig.height=3.8, fig.cap="\\textbf{Reconstruction of ontologies hierarchy.} \\textbf{(A)} Example of text associated with an ontology term from the mammalian phenotype (MP) ontology. One version includes only the term name, another version includes full-text (FT), and another has full text plus definitions of parent terms (FP+). \\textbf{(B)} Graph for a branch of the MP ontology. Dots represent ontology terms, arrows represent directional 'is a' relationships. The dark dot is the root node. \\textbf{(C)} Example of an ontology-reconstruction calculation using nearest-neighbor search. The first view shows first-ranked predictions (k=1); the second view shows the top three predictions (k=1,2,3). Labels 'hit' and 'miss' indicate when text from a given node is matched to itself via search. \\textbf{(D)} Summary of reconstruction performance averaged over the selected ontologies."}
layout(matrix(c(1,1,1,
				2,3,4,
				5,6,6,
				7,7,7),
			  ncol=3, nrow=4, byrow=T),
	   widths=c(1, 1, 1), height=c(1.1, 1.5, 0.3, .9))
#
# examples of (plain and w. parents)
example.0 = small.graph.data$"MP:0004122"$data_pos
example.1 = list("MP:0004122" = example.0[c("name", "def", "synonyms")])
example.2 = list("MP:0004122" = example.0[c("name", "def", "synonyms", "parents")])
example.2[[1]]$parents = unlist(example.2[[1]]$parents)
plot.yaml(example.2, main="Term description", max.chars=52)
lines(c(0.965, 0.975, 0.975, 0.965), c(1, 1, 0.25, 0.25), Rcssclass="yaml")
text(0.955, 0.42, "FT", Rcssclass=c("yaml", "side"))
lines(c(0.99, 1, 1, 0.99), c(1, 1, -0.01, -0.01), Rcssclass="yaml")
text(0.98, 0.11, "FT+", Rcssclass=c("yaml", "side"))
lines(c(0.94, 0.95, 0.95, 0.94), c(1, 1, 0.85, 0.85), Rcssclass="yaml")
text(0.93, 0.93, "name", Rcssclass=c("yaml", "side"))
multipanelLabel(panel.labels[1])
#
# show intended ontology graph
plot.graph(small.graph$layout, small.graph$edges,
		   show.lines="arrows", show.points="plain")
root.position = small.graph$layout["MP:0004085", ]
points(root.position[1], root.position[2], Rcssclass=c("graph", "root"))
text(root.position[1], root.position[2], "root", Rcssclass=c("graph", "root"))
multipanelLabel(panel.labels[2])
#
# show performance in mapping terms to themselves
small.results = small.graph.search.results$"mp-heartbeat.plain.data.parent"
plot.graph(small.graph$layout, small.graph$edges,
		   show.lines="simple", show.points="none",
		   main="search (k=1)", Rcssclass="bg")
plot.graph(small.graph$layout, small.results[rank<=1],
		   show.lines="arrows", show.points="self", add=TRUE,
		   Rcssclass="search")
multipanelLabel(panel.labels[3])
#
#  show performance reconstructing the graph
plot.graph(small.graph$layout, small.results[rank<=3],
		   show.lines="rank", show.points="self",
		   main="search (k=1,2,3)", Rcssclass="search")
#
par(mai=rep(0.01, 4)); plot.new();
plot.graph.legend(k=3, y=c(0.35, 0.8), Rcssclass="search")
#
# performance for finding parents
.temp = lapply(search.performance.matrix, function(x) {
	rownames(x) = gsub("plain", "FT", rownames(x))
	colnames(x) = gsub("plain", "FT", colnames(x))
	rownames(x) = gsub("parents", "FT+", rownames(x))
	colnames(x) = gsub("parents", "FT+", colnames(x))
	x
})
plot.heattable(.temp$self, .temp$parent,
			   xlab="query data", ylab="training data",
			   main1="Self retrieval", main2="Parent retrieval")
multipanelLabel(panel.labels[4])
```

Ontologies for which parent detection is better when the training is done with parents:

\footnotesize

```{r, echo=FALSE, eval=TRUE}
perf.data = search.performance.summary.wide
perf.data$train_predict = paste0(perf.data$obotype, "_", perf.data$datatype)
perf.data.wide = dcast(perf.data, id ~ train_predict,
                       value.var=c("precision_self", "precision_any_parent"))
perf.data.wide[precision_any_parent_parents_parents > precision_any_parent_plain_parents,
               c("id", "precision_any_parent_parents_parents",
				 "precision_any_parent_plain_parents")]
```

\normalsize

\clearpage


```{r fig.explaining, eval=TRUE, echo=FALSE, fig.width=6.5, fig.height=3.24, fig.cap="\\textbf{Factors affecting ontology reconstruction from text descriptions.} Panels show relations between scores from machine learning models, trained and queried with FT data, and various explanatory variables. Vertical axes show \\textbf{(A)} self-retrieval and \\textbf{(B)} parent-retrieval scores. Dots represent ontologies. Selected dots are labeled manually. "}
layout(matrix(c(1, 2, 3, 4, 5,
				6, 7, 8, 9, 10),
			  ncol=5, nrow=2, byrow=T),
	   widths=c(1.62, 1.22, 1.22, 1.22, 1.22), height=c(1.62, 1.62))
#
perf.model.data = perf.data[obotype=="plain" & datatype=="plain"]
perf.model.data$ID = toupper(perf.model.data$id)
perf.model.data$avg_chars_def = 1+perf.model.data$avg_chars_def
perf.model.data$avg_chars_comments = 1+perf.model.data$avg_chars_comments
fig.explaining(perf.model.data, "precision_self",
			   ylab="self retrieval", adj=c(0.2, -1),
			   panel.label=panel.labels[1],
			   label.ids=list("olatdv", "gno", "ncbitaxon", "lbo", "fbdv"))
fig.explaining(perf.model.data, "precision_any_parent",
			   ylab="parent retrieval", adj=c(0.2, -1),
			   panel.label=panel.labels[2],
			   label.ids=list("olatdv", "gno", c("ncbitaxon", "omo"),
							  c("lbo"), c("fbdv")))
```

Average self-precision and parent-precision

```{r}
summary(perf.model.data$precision_self)
summary(perf.model.data$precision_any_parent)
```

Lowest performing ontologies for self-retrieval

```{r}
head(perf.model.data[order(precision_self), c("id", "precision_self")], 3)
```

Worst-performing ontologies for parent-retrieval

```{r}
head(perf.model.data[order(precision_any_parent), c("id", "precision_any_parent")], 3)
```



\clearpage

```{r fig.casestudies, eval=TRUE, echo=FALSE, fig.width=6.5, fig.height=3.4, fig.cap="\\textbf{Feedback from ML for ontology development.} \\textbf{(A)} Unsupervised UMAP embedding of FBDV ontology terms (dots) according to their similarity of text-based annotations. Selected groups of terms are labeled manually. \\textbf{(B)} Parent-retrieval performance in the original FBDV ontology and a modified version with more hierarchical relations. \\textbf{(C)} Effect of diffusion on parent retrieval in the human phenotype ontology, HP. Dots represent measurements and lines are interpolations. Two series represent results on models trained with names and FT datasets; all queries performed with names data. \\textbf{(D)} Summary of errors in parent retrieval without (above) and with (below) diffusion. \\textbf{(E)} Example of an ontology term with a large parent-retrieval error. \\textbf{(F)} Activation heatmap for exemplar term representations."}
layout(matrix(c(1, 1, 3, 4, 5,
				1, 1, 3, 4, 6,
                1, 1, 7, 7, 7,
                2, 8, 8, 8, 8),
			  ncol=5, nrow=4, byrow=T),
	   widths=c(1.6, 0.8, 2.1, 0.25, 1.95), height=c(0.75, 0.85, 0.8, 1.0))
#
# FBDV ontology
#
# panel with UMAP embedding
plot_ontology_embedding(fbdv.embeddings$plain.data.plain.diff.NA.NA,
						fbdv$summary,
						show.labels=FALSE, legend.pos=c(-29, -12),
						main="FBDV (fly development)")
multipanelLabel(panel.labels[1])
fbdv.umap.tab = fbdv.embeddings$plain.data.plain.diff.NA.NA
fbdv.umap.tab = data.table(id=rownames(fbdv.umap.tab),
                           x=fbdv.umap.tab[,1], y=fbdv.umap.tab[,2])
fbdv.umap.tab = merge(fbdv.umap.tab, fbdv$terms, by="id")
fwrite(fbdv.umap.tab, file=file.path("tables", "fbdv.umap.tsv"), sep="\t")
# misc labeling
polygon(4.0 + c(-7.5, -2.5, 2.5, 7.5, 7.5, 2.5, -2.5, -7.5),
		5.0 + c(-21, -16.5, -16.5, -21, -26.5, -30, -30, -26.5),
		Rcssclass="embedding")
text(13, -21, "adulthood", Rcssclass=c("embedding", "label"))
# larval stages (e.g. FBdv:00007004, FBdv:00007005, FBdv:00005338)
polygon(c(18, 18, 26, 26), c(1, 11, 11, 1), Rcssclass="embedding")
lines(c(22, 22), c(-7, 1), Rcssclass=c("embedding", "label"))
text(12, -9, "larval stages", Rcssclass=c("embedding", "label"))
# pharate stages (e.g. )
# polygon(0.5+c(4, 4, 10, 10), c(15, 26, 26, 15), Rcssclass="embedding")
# text(11.3, 22, "pharate stages", Rcssclass=c("embedding", "label"))
# prepupal stages (e.g. FBdv:00005345, FBdv:00005342)
polygon(c(8, 8, 14, 14), c(15, 21.5, 21.5, 15), Rcssclass="embedding")
text(8, 23, "prepupal stages", Rcssclass=c("embedding", "label"))
#
# panel with fbdv variant
.fbdv.bars = modified.performance.summary[id %in% c("fbdv", "fbdv-isa") &
		obotype=="plain" & datatype=="plain"]
.fbdv.bars = setNames(.fbdv.bars$precision_any_parent, .fbdv.bars$id)
names(.fbdv.bars)[names(.fbdv.bars)=="fbdv-isa"] = "modified"
names(.fbdv.bars)[names(.fbdv.bars)=="fbdv"] = "original"
plot.simplebars(rev(.fbdv.bars), at=c(0, 0.1, 0.2, 0.3),
				main="FBDV", xlim=c(0, 0.31),
				xlab="parent-retrieval")
multipanelLabel(panel.labels[2])
#
# HP ontology
#
# line graph with diffusion settings
plot.lines(hp.search.diff.summary[difftype=="parents" & obotype=="name"],
		   x="diffvalue", y="precision_any_parent",
		   xlim=c(0, 0.5), ylim=c(0, 0.62),
           xlab="diffusion strength", ylab="parent retrieval",
		   main="HP (human phenotypes)")
add.line(hp.search.diff.summary[difftype=="parents" & obotype=="plain"],
		  x="diffvalue", y="precision_any_parent", Rcssclass=c("lines", "plain"))
legend(0.25, 0.4, c("name", "FT"), Rcssclass=c("lines"),
	   title="training data")
multipanelLabel(panel.labels[3])
# draw two histograms
hp.common.terms = intersect(hp.distances$diff$id, hp.distances$nodiff$id)
counts.nodiff = hp.distances$nodiff[id %in% hp.common.terms]$pathlen_bestN
counts.diff = hp.distances$diff[id %in% hp.common.terms]$pathlen_bestN
parplot(c(0, 1), Rcssclass=c("inthist", "tall"))
mtext(side=2, "HP terms", Rcssclass=c("inthist", "ylab"))
multipanelLabel(panel.labels[4])
plot.int.hist(counts.nodiff, main="Parent retrieval",
              ylim=c(0, 9000), xlim=c(-0.5, 12.5), Rcssclass=c("top", "nodiff"))
lines(c(3.5, 3.5, 12.5), c(1200, 2000, 2000), Rcssclass=c("inthist", "label"))
num.high.nodiff = sum(counts.nodiff >= 4)
text(8, 2900, paste0(num.high.nodiff, " terms"), Rcssclass=c("inthits", "label"))
plot.int.hist(counts.diff, main="(with diffusion)",
			  xlab="path length to parent (best of k=5)",
			  show.x.labels=TRUE,
			  ylim=c(0, 9000), xlim=c(-0.5, 12.5), Rcssclass=c("bottom", "diff"))
lines(c(3.5, 3.5, 12.5), c(1200, 2000, 2000), Rcssclass=c("inthist", "label"))
num.high.diff = sum(counts.diff >= 4)
text(8, 2900, paste0(num.high.diff, " terms"), Rcssclass=c("inthits", "label"))
#
# panel with example mapping for one hp term
plot.hp.example("HP:0020134", x2=0.52)
multipanelLabel(panel.labels[5])
#
# heatmap with features
hp.eg.ids = c("HP:0020134_name", "HP:0020134_plain",
			  "HP:0011991_plain", "HP:0012085_plain")
.m = hp.eg.matrix[, hp.eg.ids]
colnames(.m) = gsub("plain", "FT", colnames(.m))
maxval = function(x) {
	suppressWarnings(max(x, na.rm=TRUE))
}
maxval.above = function(x, min.value) {
	maxval(x) > min.value
}
.m = .m[order(-apply(.m, 1, maxval)), , drop=FALSE]
plot.heatmap(.m[apply(.m, 1, maxval.above, 0.10), ,drop=FALSE],
             color.max=0.5, main="Activated features")
text(-1.0, -6.8, "ML features:", xpd=1, cex=1, adj=c(1, 0.5),
	 Rcssclass=c("heatmap", "xlab"))
multipanelLabel(panel.labels[6])
```

Number of terms in fbdv: `r ontologies.summary[id=="fbdv"]$num_terms`

Number of terms in hp: `r ontologies.summary[id=="hp"]$num_terms`

Depth information:

```{r}
field.summary.stats[id %in% c("fbdv", "hp") & field=="depth"]
```

Parent-retrieval in modified ontologies

```{r}
modified.performance.summary[obotype=="plain" & datatype=="plain"]
```

Performance in hp parent retrieval before and after diffusion

```{r}
hp.search.diff.summary[abs(diffvalue-0.0)<0.01 & obotype=="plain" & difftype=="parents"]
hp.search.diff.summary[abs(diffvalue-0.3)<0.01 & obotype=="plain" & difftype=="parents"]
```

Path length for HP example

```{r}
hp.distances$nodiff[id=="HP:0020134",]
```

